---
title: "Project 3"
author: "Author: Volha Tryputsen"
output:
  html_document:
    highlight: haddock
    theme: cosmo
    toc: yes
    toc_depth: 6
    toc_float:
      collapsed: yes
      smooth_scroll: no
  pdf_document:
    toc: yes
    toc_depth: '6'
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```


```{r libraries}
library(tidyverse)
library(readr) # read the data
library(janitor)
library(dplyr)
library(DataExplorer) 
#https://towardsdatascience.com/simple-fast-exploratory-data-analysis-in-r-with-dataexplorer-package-e055348d9619
library(knitr)
library(lme4)
library(gee)
library(nlme)
library(gridExtra)
library(kableExtra)
library(dotwhisker) #https://www.rdocumentation.org/packages/dotwhisker/versions/0.5.0/topics/dwplot
```



# Data  
```{r}
datt = 
  read_csv("data.csv") %>% 
  clean_names()

dat = 
  datt %>%
  rename(treatment = treatment_group, id = subject_id) %>%
  mutate(treatment = recode(treatment, A = "placebo", B = "low_dose", C = "high_dose")) %>%
  mutate(treatment = as.factor(treatment), 
         id = as.factor(id),
         gender = as.factor(gender),
        treatment = forcats::fct_relevel(treatment, "placebo", "low_dose", "high_dose"))
 
 
head(dat) 
```

# Objective  
The primary objective of the study was to assess whether DAR-0100A treatment at low (Treatment_group = B in
associated data) or high dose (Treatment_group = C in associated data) compared to placebo (Treatment_group = A in associated data) is able to improve memory as measured by the mem_comp rating.   
Researchers were specifically interested in:  

- the treatment effects at day 5, 19, and 90 and 
- whether the treatment effects differed over time  


# EDA  
## Sample size  

total of 47 subjects.  
```{r}
dat %>%
  mutate(day = as.factor(day)) %>%
  group_by(treatment, day) %>% 
  dplyr::summarize(n = n()) %>%
  spread(key = day, value = n) %>%
        kable(., caption = "Table 1", digits = 0) %>%
        kableExtra::kable_styling(bootstrap_options = c("striped", "hover",  "condensed", "responsive"), 
                              full_width = F, position = "left")
```


## Separate demographics  
```{r}
dat_dem = 
  dat %>%
  select(c(id, age, gender, treatment)) %>%
  distinct()
  
# also has BL scores
dat_dem_bl = 
  dat %>%
  filter(day == 0) %>%
  mutate(mem_comp_bl = mem_comp) %>%
  dplyr::select(-c(day, mem_comp))

```

### Age distribution 
```{r}
dat_dem_bl %>%
  ggplot(aes(x = treatment, y = age, fill = treatment)) +
  geom_boxplot() +
  theme_bw()

summary(aov(age ~ treatment, data = dat_dem_bl)) # no differences
```

### Gender 
```{r}
ftable(dat_dem_bl$treatment, dat_dem_bl$gender) 

chisq.test(ftable(dat_dem_bl$treatment, dat_dem_bl$gender))
```





## Missing values summary  
```{r}
# table with the number of visits
dat_miss = 
  dat %>%
  group_by(id) %>%
  summarize(n = n()) %>%
  ungroup() 

# join with demographics data
dat_miss = 
  dat_miss %>%
  left_join(., y = dat_dem, by = "id") %>%
  distinct()

# summary statistics for the number of follow up visits  
dat_miss %>%
  group_by(treatment) %>%
  summarise(mean(n), sd(n), max(n), min(n)) %>%
  kable(., caption = "Table 2", digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover",  "condensed", "responsive"), 
                              full_width = F, position = "left")

# barchart of count for each treatment group
dat_miss %>%
  ggplot(aes(x = n, fill= treatment))+
  geom_bar()+
  theme_bw() +
  facet_grid(~treatment)
```

code missing values with NA`s  
```{r}
dat_all = 
  dat %>%
  spread(key = day, value = mem_comp) %>%
  gather(key = day, value = mem_comp, `0`:`90`) %>%
  mutate(day = as.numeric(day))


dat_na = 
  dat_all %>%
  group_by(id) %>%
  summarize(n = sum(is.na(mem_comp))) %>%
  filter(n>0) %>%
  ungroup()

dat_na = 
  dat_na %>%
  left_join(y = dat_miss, by = "id") %>%
  distinct()


dat_na %>%
  mutate(n = as.factor(n.x)) %>%
  ggplot(aes(x = id, y = n.x, color = treatment)) +
  geom_point() +
  facet_grid(~treatment) +
  theme_bw() 

dim(dat_na)
```



## Spaghetti plot  
### Individual  
```{r}
dat %>%
  ggplot(aes(x = day, y = mem_comp, colour = id)) +
  geom_line() +
  geom_point(alpha = 0.3, size = 1) +
  facet_grid(~treatment) +
  theme_bw() +
  theme(legend.position = "none") 
```


### Mean  
```{r}
dat %>%
  as_tibble() %>%
  group_by(treatment, day) %>% 
  dplyr::summarise(n=n(), mean = mean(mem_comp), sd = sd(mem_comp)) %>%
  mutate(se = sd/sqrt(n), mean = round(mean, 2)) %>%
  ggplot(aes(x = day, y = mean, label = n, colour = treatment)) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width=1) +
  geom_line() + 
  geom_point() + 
  theme_bw() 
```


## Baseline  

```{r}
# check differenes in baseline between treatmet groups
dat_dem_bl %>%
  ggplot(aes(x = treatment, y = mem_comp_bl, fill = treatment)) +
  geom_boxplot() +
  theme_bw()

# test for meme_comp differences at baseline
summary(aov(mem_comp_bl ~ treatment, data = dat_dem_bl)) # no differences
```


Calculate differences from baseline  
```{r}
dat_diff = 
  dat %>%
  left_join(., y = dat_dem_bl[,c("id", "mem_comp_bl")], by = "id") %>%
  mutate(mem_comp_diff = mem_comp - mem_comp_bl) #%>%
  #filter(day %in% c(5,19,90))
```


Spaghheti plot of differences   
```{r}
dat_diff %>%
  ggplot(aes(x = day, y = mem_comp_diff, colour = id)) +
  geom_line() +
  geom_point(alpha = 0.3, size = 1) +
  facet_grid(~treatment) +
  theme_bw() +
  theme(legend.position = "none") 
```


Mean spaghetti plots of differences  
```{r}
dat_diff %>%
  as_tibble() %>%
  group_by(treatment, day) %>% 
  dplyr::summarise(n=n(), mean = mean(mem_comp_diff), sd = sd(mem_comp_diff)) %>%
  mutate(se = sd/sqrt(n), mean = round(mean, 2)) %>%
  ggplot(aes(x = day, y = mean, label = n, colour = treatment)) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width=1) +
  geom_line() + 
  geom_point() + 
  theme_bw() 
```





## Check MCAR 
```{r}
# if observation is missing- asign 0, if observation was observed - assign 1
dat_all = 
  dat_all %>%
  mutate(missing = ifelse(is.na(mem_comp), 0, 1),
         missing = as.factor(missing))


dat_all

##fit the logistic regression with random effect 
m1 <- glmer(missing ~ age + gender + day + treatment + (1|id), data = dat_all, family = binomial)
m1_sum = summary(m1)


m1_out = round(cbind(estimate = m1_sum$coefficients[,1], SE = m1_sum$coefficients[,2], 
      p_value = m1_sum$coefficients[,4], exp_est = exp(m1_sum$coefficients[,1]), 
      exp_est_lb =  exp(m1_sum$coefficients[,1] - 1.96*m1_sum$coefficients[,2]),
      exp_est_ub =  exp(m1_sum$coefficients[,1] + 1.96*m1_sum$coefficients[,2])), 4)[-1,]

m1_out %>%
kable(., digits = 3) %>%
        kableExtra::kable_styling(bootstrap_options = c("striped", "hover",  "condensed", "responsive"), 
                              full_width = F, position = "left")
```

Day is significant, therefore MCAR is not true.  


## MAR and LME    
Thus, ussume MAR and use conditional longitudinal model approach.  

### (m1) mem_comp ~ predictors  
Fit random intercept models. Time is a factor.  

```{r}
datf = 
  dat %>%
  mutate(day = as.factor(day))

#random intercept
lme1 <- lme(mem_comp ~ day*treatment + age + gender, random=~1|id, data = datf)
AIC(lme1)

lme1_ar1 = try(update(lme1, corr = corAR1()))
lme1_comp_symm = try(update(lme1, corr = corCompSymm()))
AIC(lme1, lme1_ar1, lme1_comp_symm)
BIC(lme1, lme1_ar1, lme1_comp_symm)


summary(lme1)$tTable
```

Diagnostics
```{r }
model = lme1
lme_plot <- data.frame(datf,
                       residuals = residuals(model),
                        fitted = fitted(model),
                        observed = datf$mem_comp)

## Residual Histogram
ggplot(lme_plot, aes(x = residuals)) +
  geom_histogram(bins = 40, aes(y = ..density..),
                 fill = "black", color = "white") +
  stat_function(fun = dnorm, 
                args = list(mean = mean(lme_plot$residuals),
                            sd = sd(lme_plot$residuals)), col = "red",linetype = "dashed") +
  theme_bw() +
  xlab("Residuals") +
  ylab("Density") +
  ggtitle("Histogram of Residuals") 



## Residual Q-Q Plot
  y <- quantile(lme_plot$residuals, c(0.25, 0.75)) # Find the 1st and 3rd quartiles
  x <- qnorm( c(0.25, 0.75))     # Find the matching normal values on the x-axis
  slope <- diff(y) / diff(x)     # Compute the line slope
  int <- y[1] - slope * x[1]     # Compute the line intercept
ggplot(lme_plot, aes(sample = residuals)) +
  stat_qq(distribution = qnorm) +
  geom_abline(intercept = int, slope = slope, color = "red", linetype = "dashed") + 
  theme_bw() +
  ggtitle("Q-Q Plot of Residuals") +
  xlab("Theoretical") +
  ylab("Sample") 

  
  ## Fitted vs. Residuals Scatter Plot
ggplot(data = lme_plot, aes(x = fitted, y = residuals)) +
    geom_point() +
    geom_smooth(method = "loess") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    ggtitle("Fitted vs. Residuals") +
  theme_bw() +
    xlab("Fitted") +
    ylab("Residuals") +
    guides(color = FALSE)

## Fitted vs. Observed Scatter Plot
ggplot(data = lme_plot, aes(x = fitted, y = observed, color = treatment))+
  geom_point() +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
    ggtitle("Fitted vs. Observed") +
    theme(legend.position="none") +
  theme_bw() +
    xlab("Fitted") +
    ylab("Observed") 

## Random effect
qqnorm(lme1, ~ranef(.))
```

Final model:    
```{r}
my_model = summary(lme1)$tTable
my_model %>%
      kable(., caption = "Table 3", digits = 4) %>%
      kableExtra::kable_styling(bootstrap_options = c("striped", "hover",  "condensed", "responsive"), 
                              full_width = F, position = "left")
```



### (m2) mem_comp ~ baseline + predictors  ***  

```{r}
dat_m2 = 
  dat_diff %>%
  mutate(day = as.factor(day)) %>%
  filter(day %in% c(5,19,90))

#random intercept
lme2 <- lme(mem_comp ~ mem_comp_bl + day*treatment + age + gender, random=~1|id, data = dat_m2)
AIC(lme2)

lme2_ar1 = try(update(lme2, corr = corAR1()))
lme2_comp_symm = try(update(lme2, corr = corCompSymm()))
AIC(lme2, lme2_ar1, lme2_comp_symm)
BIC(lme2, lme2_ar1, lme2_comp_symm)


# final model is random intercept with uncorrelated residuals and random effects
lme2_sum = summary(lme2)
# only treatment with high dose is different from placebo at day 5 and day 19, compaed to BL

lme2_out = round(cbind(estimate = lme2_sum$tTab[,1], SE = lme2_sum$tTab[,2], 
                     p_value = lme2_sum$tTab[,5], 
                     est_lb =  lme2_sum$tTab[,1] - 1.96*lme2_sum$tTab[,2],
                     est_ub =  lme2_sum$tTab[,1] + 1.96*lme2_sum$tTab[,2]), 4)
lme2_out %>%
  kable(., digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover",  "condensed", "responsive"), 
                              full_width = F, position = "left")
```


### (m3) mem_comp_diff ~ predictors  

```{r}
#random intercept
lme3 <- lme(mem_comp_diff ~ day*treatment + age + gender, random=~1|id, data = dat_m2)
AIC(lme3)

lme3_ar1 = try(update(lme3, corr = corAR1()))
lme3_comp_symm = try(update(lme3, corr = corCompSymm()))
AIC(lme3, lme3_ar1, lme3_comp_symm)
BIC(lme3, lme3_ar1, lme3_comp_symm)


# final model is random intercept with uncorrelated residuals and random effects
round(summary(lme3)$tTable,3)
# only treatment with high dose is different from placebo at day 5 and day 19, compaed to BL

```



### (m4) mem_comp_diff ~ baseine + predictors  

```{r}
#random intercept
lme4 <- lme(mem_comp_diff ~ mem_comp_bl + day*treatment + age + gender, random=~1|id, data = dat_m2)
AIC(lme4)

lme4_ar1 = try(update(lme4, corr = corAR1()))
lme4_comp_symm = try(update(lme4, corr = corCompSymm()))
AIC(lme4, lme4_ar1, lme4_comp_symm)
BIC(lme4, lme4_ar1, lme4_comp_symm)


# final model is random intercept with uncorrelated residuals and random effects
round(summary(lme4)$tTable,3)
# only treatment with high dose is different from placebo at day 5 and day 19, compaed to BL

```


### (m5)
```{r}
dat_m5 = 
  dat_diff %>%
  filter(day %in% c(5,19,90))

#random intercept
lme5 <- lme(mem_comp ~ mem_comp_bl + day*treatment + age + gender, random=~1|id, data = dat_m5)
AIC(lme5)

round(summary(lme5)$tTable,3)
```


Diagnostics
```{r }
model = lme5
lme_plot <- data.frame(dat_m5,
                       residuals = residuals(model),
                        fitted = fitted(model),
                        observed = dat_m5$mem_comp)

## Residual Histogram
ggplot(lme_plot, aes(x = residuals)) +
  geom_histogram(bins = 40, aes(y = ..density..),
                 fill = "black", color = "white") +
  stat_function(fun = dnorm, 
                args = list(mean = mean(lme_plot$residuals),
                            sd = sd(lme_plot$residuals)), col = "red",linetype = "dashed") +
  theme_bw() +
  xlab("Residuals") +
  ylab("Density") +
  ggtitle("Histogram of Residuals") 



## Residual Q-Q Plot
  y <- quantile(lme_plot$residuals, c(0.25, 0.75)) # Find the 1st and 3rd quartiles
  x <- qnorm( c(0.25, 0.75))     # Find the matching normal values on the x-axis
  slope <- diff(y) / diff(x)     # Compute the line slope
  int <- y[1] - slope * x[1]     # Compute the line intercept
ggplot(lme_plot, aes(sample = residuals)) +
  stat_qq(distribution = qnorm) +
  geom_abline(intercept = int, slope = slope, color = "red", linetype = "dashed") + 
  theme_bw() +
  ggtitle("Q-Q Plot of Residuals") +
  xlab("Theoretical") +
  ylab("Sample") 

  
  ## Fitted vs. Residuals Scatter Plot
ggplot(data = lme_plot, aes(x = fitted, y = residuals)) +
    geom_point() +
    geom_smooth(method = "loess") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    ggtitle("Fitted vs. Residuals") +
  theme_bw() +
    xlab("Fitted") +
    ylab("Residuals") +
    guides(color = FALSE)

## Fitted vs. Observed Scatter Plot
ggplot(data = lme_plot, aes(x = fitted, y = observed, color = treatment))+
  geom_point() +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
    ggtitle("Fitted vs. Observed") +
    theme(legend.position="none") +
  theme_bw() +
    xlab("Fitted") +
    ylab("Observed") 


qqnorm(lme5, ~ranef(.))
```




## MAR mice   

```{r}
library(mice)
library(reshape)

dat_sh = 
  dat %>%
  spread(key = day, value = mem_comp) %>%
  select(-c(age, gender)) %>%
  dplyr::rename(day_0 = `0`, day_5 = `5`, day_19 = `19`, day_90 = `90`) %>%
  as.data.frame()
head(dat_sh)


md.pattern(dat_sh)


# impute
m.out <- mice(dat_sh, seed = 9, m = 20, pred=quickpred(dat_sh, mincor=0, exclude=c("id", "treatment")))


# complete
imp.data <- as.list(1:20)
for(i in 1:20){
  imp.data[[i]] <- complete(m.out, action=i)
}

imp.data[[1]] %>% head()

# reshape
imp.data <- lapply(imp.data, melt, id=c("id","treatment", "day_0"))


f_clean <- function(x){
  x %>%
  separate(variable, into = c("out", "day")) %>%
  select(-out) %>%
  dplyr::rename(mem_comp_imp = value) %>%
  left_join(., y = dat_dem_bl[,c("id", "age", "gender")], by = "id")
} 

imp.data1 <- lapply(imp.data, f_clean)
imp.data1[[1]] %>% head()


# analyse
#imp.fit <- lapply(imp.data1, FUN=function(x){
#  summary(lme(mem_comp_imp ~ day_0 + day*treatment + age + gender, random=~1|id, data = x))$tTab
#  #lme(value ~ as.numeric(variable)+(1|id), data=x) 
#})

imp.fit <- lapply(imp.data1, FUN=function(x){
  lme(mem_comp_imp ~ day_0 + day*treatment + age + gender, random=~1|id, data = x)
  #lme(value ~ as.numeric(variable)+(1|id), data=x) 
})

pool_sum = summary(pool(imp.fit))

pool_out = round(cbind(estimate = pool_sum[,1], SE = pool_sum[,2], 
                     p_value = pool_sum[,5], 
                     est_lb =  pool_sum[,1] - 1.96*pool_sum[,2],
                     est_ub =  pool_sum[,1] + 1.96*pool_sum[,2]), 4)
rownames(pool_out) <- c(rownames(pool_sum))

pool_out %>%
  kable(., digits = 3) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover",  "condensed", "responsive"), 
                              full_width = F, position = "left")

```

### imputed data
#### 1
```{r eval=FALSE}
d1 = imp.data[[1]]
head(d1)
d1 = d1 %>% 
  separate(variable, into = c("out", "day")) %>%
  select(-out) %>%
  mutate(day = as.numeric(day),
         mem_comp = value)


d1 %>%
  ggplot(aes(x = day, y = mem_comp, colour = id)) +
  geom_line() +
  geom_point(alpha = 0.3, size = 1) +
  facet_grid(~treatment) +
  theme_bw() +
  theme(legend.position = "none") 

d1 %>%
  as_tibble() %>%
  group_by(treatment, day) %>% 
  dplyr::summarise(n=n(), mean = mean(mem_comp), sd = sd(mem_comp)) %>%
  mutate(se = sd/sqrt(n), mean = round(mean, 2)) %>%
  ggplot(aes(x = day, y = mean, label = n, colour = treatment)) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width=1) +
  geom_line() + 
  geom_point() + 
  theme_bw() 
```

